### 2025-08-09 draft.qmd
20250809 - lineup optimization

```{r}
library("tidyverse")
```

@@

load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/player_stats_seasonal.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/player_stats_weekly.RData", fsep = ""))
```

```{r}
load(file = "./data/player_stats_seasonal.RData")
load(file = "./data/player_stats_weekly.RData")
```

@@

@@

@@

>>>>

```{r}
library("tidyverse")
library("Rglpk")
```

@@

load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/player_stats_seasonal.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/player_stats_weekly.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/players_projections_seasonal.RData", fsep = ""))
```

```{r}
load(file = "./data/player_stats_seasonal.RData")
load(file = "./data/player_stats_weekly.RData")
load(file = "./data/players_projections_seasonal.RData")
```

@@

FantasyFootballAnalytics.net provides tools to help you win your [snake draft](#sec-draftSnake): <https://fantasyfootballanalytics.net/win-your-snake-draft>.
For draft tools for [snake drafts](#sec-draftSnake), see the Fantasy Football Analytics web application: <https://apps.fantasyfootballanalytics.net>.
The web application allows users to specify their custom league settings to calculate projected points that are customized to their league and helps users select which players to draft that maximize the players' [value over replacement players](#sec-fantasyValueVORP) (VORP).

@@

FantasyFootballAnalytics.net provides tools to help you win your [auction draft](#sec-draftSnake) (<https://fantasyfootballanalytics.net/win-your-snake-draft>) and daily fantasy sports (DFS) leagues (<https://fantasyfootballanalytics.net/how-to/dfsoptimizerhelp>).
For draft and player selection tools for [auction drafts](#sec-draftSnake) and DFS leagues, see the Fantasy Football Analytics web application: <https://apps.fantasyfootballanalytics.net>.
The web application allows users to specify their custom league settings to calculate projected points that are customized to their league, and to identify the optimal lineup given their league settings and salary cap.

@@

### Lineup Optimization {#sec-lineupOptimization}

For [auction drafts](#sec-draftAuction) and daily fantasy sports (DFS) leagues, a key goal is to select the best lineup or team within the given cap.
For instance, in an [auction drafts](#sec-draftAuction), you may have a \$200 cap and your goal is to draft the best team that you can with the \$200.
Using a mixed-integer programming solver, we can identify the lineup with the most projected points given a particular salary cap.
For projected points for a player, we use their average projected points across all available projection sources.
For a player's estimated cost, we use their [Average Auction Value](#sec-fantasyValueADPAAV) (AAV), rounded up to the nearest integer.

We first subset and prepare the data so we can perform the lineup optimization.

```{r}
mostRecentProjections <- players_projections_seasonal_average_merged %>% 
  filter(season == max(season, na.rm = TRUE)) %>% 
  filter(pos %in% c("QB","RB","WR","TE","K","DST")) %>% 
  filter(avg_type == "average")

# Convert NAs for fantasy points to zero
mostRecentProjections$points[which(is.na(mostRecentProjections$points))] <- 0

# Convert NA for AAV to $1
mostRecentProjections$aav[which(is.na(mostRecentProjections$aav))] <- 1

# Round up AAV to nearest integer
mostRecentProjections$cost <- ceiling(mostRecentProjections$aav)

mostRecentProjections$name <- paste(mostRecentProjections$first_name, mostRecentProjections$last_name)
```

Next, we set up the constraints.
Our example is adapted from <https://stackoverflow.com/questions/15147398/optimize-value-with-linear-or-non-linear-constraints-in-r>.
Our constraints were:

1) 9 total players of the following number at each of following positions:
- 1 Quarterback
- 2 Running Backs
- 2 Wide Receivers
- 1 Tight End
- 1 Flex: RB/WR/TE
- 1 K
- 1 DST

and 2) salary cap of \$200.

```{r}
# number of players
num.players <- nrow(mostRecentProjections)

# 1 for every player
total_players <- rep(1, num.players)

# objective:
#f <- pts

# the variable are booleans
var.types <- rep("B", num.players)

# the constraints

A <- rbind(
  as.numeric(mostRecentProjections$pos == "QB"),                # num QB
  as.numeric(mostRecentProjections$pos == "RB"),                # num RB
  as.numeric(mostRecentProjections$pos == "WR"),                # num WR
  as.numeric(mostRecentProjections$pos == "TE"),                # num TE
  as.numeric(mostRecentProjections$pos == "K"),                 # num K
  as.numeric(mostRecentProjections$pos == "DST"),               # num DST
  as.numeric(mostRecentProjections$pos %in% c("RB","WR","TE")), # num flex
  total_players,
  mostRecentProjections$cost)                                   # player cost

dir <- c(
  "==",
  ">=",
  ">=",
  ">=",
  "==",
  "==",
  "==",
  "==",
  "<=")

b <- c(
  1,   # 1 QB
  2,   # at least 2 RBs
  2,   # at least 2 WRs
  1,   # at least 1 TE
  1,   # 1 K
  1,   # 1 DST
  6,   # 1 RB/WR/TE flex (i.e., 6 total RB/WR/TEs)
  9,   # 9 total players
  200) # salary cap of $200
```

The optimizer thus seeks to identify the lineup with the most projected points that meets the above constraints (within a salary cap of $200).
We perform the optimization using the `Rglpk::Rglpk_solve_LP()` function of the `Rglpk` package [@R-Rglpk].

```{r}
sol <- Rglpk::Rglpk_solve_LP(
  obj = mostRecentProjections$points,
  mat = A,
  dir = dir,
  rhs = b,
  types = var.types,
  max = TRUE)
```

Here is the solution obtained by the solver:

```{r}
sol
```

A zero for `status` indicates that it found a solution:

```{r}
sol$status
```

Here is the maximum value obtained in terms of the sum of projected points for the optimal lineup:

```{r}
sol$optimum
```

Here is a vector indicating whether or not a given player was selected for the optimal lineup (0 = no; 1 = yes):

```{r}
sol$solution
```

Here is the optimal lineup that maximized the sum of projected points within the salary cap:

```{r}
mostRecentProjections[which(sol$solution == 1), c("name","pos","points")]
```

For a web application that includes a lineup optimizer for selecting the optimal lineup for [auction drafts](#sec-draftAuction) and for DFS leagues, see the Fantasy Football Analytics web application: <https://apps.fantasyfootballanalytics.net>.
The web application allows users to specify their custom league settings, select players to draft or exclude, and exclude risky players, among other features.

### 2025-08-09 index.qmd
20250805 - acknowledgment
20250805 - fix
20250806 - robust load
20250809 - lineup optimization

@@

# Automatically create a bib database for R packages
packagesForBibliography <- c(
  "petersenlab", "ffanalytics",
  "knitr", "rmarkdown", "broom", "broom.mixed", "psych", "downlit", "xml2", 
  "gsisdecoder", "progressr", "DescTools", "pwr", "pwrss", "WebPower", "XICOR", 
  "dagitty", "ggdag", "ggtext", "gghighlight", "ggExtra", "grid", "patchwork", 
  "pROC",  "lme4", "lmerTest", "MuMIn", "emmeans", "pbkrtest", "sjstats", 
  "AICcmodavg", "rstan", "brms", "tidybayes", "bbmle", "fitdistrplus", "sn", 
  "mclust", "magrittr", "viridis", "viridisLite", "msir", "plotly", "webshot2", 
  "quantmod", "fPortfolio", "NMOF", "nFactors", "xts", "zoo", "forecast",
  "stringi", "parallelly", "doParallel", "missRanger", "ggridges", "powerjoin", 
  "caret", "LongituRF", "gpboost", "cmdstanr", "mgcv", "rms", "car", "lavaan", 
  "lavaanPlot", "mice", "miceadds", "interactions", "robustbase", "ordinal", 
  "MASS", "data.table", "future", "future.apply", "SimDesign", "domir", 
  "GGally",
  "tidyverse", "ggplot2", "dplyr", "lubridate", # packages that are part of the tidyverse ecosystem
  "nflverse", "nflfastR", "nflseedR", "nfl4th", "nflreadr", "nflplotR", # packages that are part of the nflverse ecosystem
  "tidymodels", "rsample", "parsnip", "recipes", "tune", "yardstick", # packages that are part of the tidymodels ecosysyem
  "easystats", "see", "performance", "parameters", "correlation", "effectsize", # packages that are part of the easystats ecosystem
  "learnr", "remotes", "shiny", "PearsonDS", "lavaangui", "dynamic" # packages that are cited but not used for computation (and thus do not need to be installed by me or readers)
)

@@

This work would not be possible without the important contributions of many people.
This book was made possible by the hard work of the package developers of the `nflverse` ecosystem of R packages, who made football data available, including Sebastian Carl, Ben Baldwin, Lee Sharpe, Tan Ho, and John Edwards.
I thank Dr. Benjamin Motz, who provided consultation and many helpful resources based on his fantasy football statistics class.
I also thank key members of [FantasyFootballAnalytics.net](http://fantasyfootballanalytics.net), including Val Pinskiy, Andrew Tungate, Dennis Andersen, Adam Peterson, Jesse Kartes, and Chris Russo, who helped develop and provide fantasy football-related resources and who helped sharpen my thinking about the topic.
In particular, Andrew Tungate and Dennis Andersen helped develop the [`ffanalytics` `R` package](https://github.com/FantasyFootballAnalytics/ffanalytics) [@R-ffanalytics], which scrapes fantasy football projections from the web and calculates projections averaged across sources.
Andrew Tungate also made important additions to the [`ffanalytics` `R` package](https://github.com/FantasyFootballAnalytics/ffanalytics) [@R-ffanalytics] to allow computing fantasy points given projected (or actual) player statistics, which was necessary for conducting a [bootstrap simulation](#sec-simulationProjectedPointsBootstrapping) of players' projected fantasy points.
I am fortunate to work with such a talented, passionate, and creative team.
I also thank Professor Patrick Carroll, who taught me the value of statistics for answering important questions.

>>>>

# Objects loaded but not generated during GitHub Action run: betaRegression.RData, bayesianMixedModelFit.RData, modelPredictions_bayesian.RData, monteCarloSim_results.RData

@@

# Automatically create a bib database for R packages
packagesForBibliography <- c(
  "petersenlab", "ffanalytics",
  "knitr", "rmarkdown", "broom", "broom.mixed", "psych", "downlit", "xml2", 
  "gsisdecoder", "progressr", "DescTools", "pwr", "pwrss", "WebPower", "XICOR", 
  "dagitty", "ggdag", "ggtext", "gghighlight", "ggExtra", "grid", "patchwork", 
  "pROC",  "lme4", "lmerTest", "MuMIn", "emmeans", "pbkrtest", "sjstats", 
  "AICcmodavg", "rstan", "brms", "tidybayes", "bbmle", "fitdistrplus", "sn", 
  "mclust", "magrittr", "viridis", "viridisLite", "msir", "plotly", "webshot2", 
  "quantmod", "fPortfolio", "NMOF", "nFactors", "xts", "zoo", "forecast",
  "stringi", "parallelly", "doParallel", "missRanger", "ggridges", "powerjoin", 
  "caret", "LongituRF", "gpboost", "cmdstanr", "mgcv", "rms", "car", "lavaan", 
  "lavaanPlot", "mice", "miceadds", "interactions", "robustbase", "ordinal", 
  "MASS", "data.table", "future", "future.apply", "SimDesign", "domir", 
  "GGally","Rglpk",
  "tidyverse", "ggplot2", "dplyr", "lubridate", # packages that are part of the tidyverse ecosystem
  "nflverse", "nflfastR", "nflseedR", "nfl4th", "nflreadr", "nflplotR", # packages that are part of the nflverse ecosystem
  "tidymodels", "rsample", "parsnip", "recipes", "tune", "yardstick", # packages that are part of the tidymodels ecosysyem
  "easystats", "see", "performance", "parameters", "correlation", "effectsize", # packages that are part of the easystats ecosystem
  "learnr", "remotes", "shiny", "PearsonDS", "lavaangui", "dynamic" # packages that are cited but not used for computation (and thus do not need to be installed by me or readers)
)

@@

This work would not be possible without the important contributions of many people.
This book was made possible by the hard work of the package developers of the `nflverse` ecosystem of R packages, who made football data available, including Sebastian Carl, Ben Baldwin, Lee Sharpe, Tan Ho, and John Edwards.
I thank Dr. Benjamin Motz, who provided consultation and many helpful resources based on his fantasy football statistics class.
I also thank key members of [FantasyFootballAnalytics.net](http://fantasyfootballanalytics.net), including Val Pinskiy, Andrew Tungate, Dennis Andersen, Adam Peterson, Jesse Kartes, and Chris Russo, who helped develop and provide fantasy football-related resources and who helped sharpen my thinking about the topic.
In particular, Andrew Tungate and Dennis Andersen helped develop the [`ffanalytics` `R` package](https://github.com/FantasyFootballAnalytics/ffanalytics) [@R-ffanalytics], which scrapes fantasy football projections from the web and calculates projections averaged across sources.
Andrew Tungate also made important additions to the [`ffanalytics` `R` package](https://github.com/FantasyFootballAnalytics/ffanalytics) [@R-ffanalytics] to allow computing fantasy points given projected (or actual) player statistics, which was necessary for conducting a [bootstrap simulation](#sec-simulationProjectedPointsBootstrapping) of players' projected fantasy points.
I am fortunate to work with such a talented, passionate, and creative team.
I also thank Professor Patrick Carroll, who taught me the value of statistics for answering important questions.
Finally, I thank my wife who, in addition to designing the book cover, is a constant source of love and support.
